---
title: "Lab 14 : Taxonomy visualization with Sankey plots"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: united
    highlight: tango
---

## Learning objectives
* How to visualize taxonomies using a Sankey plot

## Libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(zoo)
library(knitr)
```

## Overview

By now you have realized that it is difficult to make a single graph that shows that taxonomic levels of your study group. Sankey plots have become popular in the last few years as a way to visual several or all taxonomic levels in a single graph. Today we will make Sankey plots using a R Shiny app developed by Florian Breitwieser. First we will need to use R to make the proper input files. 

## NEON MAG table

### Base level dataframes with the NEON MAGs

```{r, warning=FALSE, message=FALSE}
NEON_MAGs <- read_csv("data/NEON/GOLD_Study_ID_Gs0161344_NEON_edArchaea.csv") %>% 
  # remove columns that are not needed for data analysis
  select(-c(`GOLD Study ID`, `Bin Methods`, `Created By`, `Date Added`)) %>% 
  # create a new column with the Assembly Type
  mutate("Assembly Type" = case_when(`Genome Name` == "NEON combined assembly" ~ `Genome Name`,
                            TRUE ~ "Individual")) %>% 
  mutate_at("Assembly Type", str_replace, "NEON combined assembly", "Combined") %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus"), "; ", remove = FALSE) %>% 
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "S-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

### Filter for individual assemblies
```{r}
NEON_MAGs_ind <- NEON_MAGs %>% 
  filter(`Assembly Type` == "Individual") 
```

### Filter for combined assemblies
```{r}
NEON_MAGs_co <- NEON_MAGs %>% 
  filter(`Assembly Type` == "Combined") 
```


## Make input for for Pavian Sankey Plot (from IMG format for GTDB taxonomy

### Make data frame in Pavian Sankey format
```{r}
# Select the GTDB Taxonomic lineage and separate into taxonomic levels
sankey_data <- NEON_MAGs_co %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  # NAs are likely Archaea
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  # Pavian format requires p__ etc
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), "; ") 

# Fill in the NAs with the taxonomic the higher taxonomic level to the left
sankey_data[] <- t(apply(sankey_data, 1, zoo::na.locf))

# Put the data into a format that can be read by the Sankey App

sankey_data <- sankey_data %>% 
  unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")  

# Create format for Pavian with counts for each taxonomic level
sankey_data_s <- sankey_data
sankey_data_g <- sankey_data
sankey_data_f <- sankey_data
sankey_data_o <- sankey_data
sankey_data_c <- sankey_data
sankey_data_p <- sankey_data
sankey_data_d <- sankey_data

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)  

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 
# rename for Pavian format
  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

# Write file to input to Pavian Sankey
write_tsv(sankey_data_allTaxa, "data/NEON/NEON_MAG_co_pavian.txt")
```


## Pavian Sankey Plot

### Generating the base Sankey graph
* Upload the above pavian formatted file to [Pavian Shiny](https://fbreitwieser.shinyapps.io/pavian/)
* Click Generate html report
* Click on Sample then Configure Sankey to see the plot

### Parameters to adjust on the Sankey graph
* Taxonomic ranks D,P,C,O,F and maybe G
* Increase scale between nodes to 1.2-1.4
* Increase Figure height to 500-1000
* Increase taxa to max level which still conveys key points and makes a good visual 

### Saving the html file and/or save the html file as a png.
* Save network

* Use the same parameters for both the individual and combined assemblies

### Results files for all NEON taxa

This code will `embed` the Sankey html results files within the Rmarkdown document.  You should use it when making your report.

```{r, fig.width=10}
knitr::include_url("images/lab14/sankey-NEON_MAG_ind_pavian.txt.html")
knitr::include_url("images/lab14/sankey-NEON_MAG_co_pavian.txt.html")
```

## Exercises

### Exercise 1

Make a Sankey plot for the individual assemblies for your Phylum at all sites.

### Exercise 2

Make a Sankey plot for the combined assemblies for your Phylum.

### Exercise 3

Make a Sankey plot for all taxa at your site.



